---
title: "Victim Count in School vs. Non-School Shootings"
author: "Robyn Boyland"
description: "Do police respond better when victims are younger?"
output:
  distill::distill_article:
    self_contained: false
---



## Introduction

My original interest was to explore data related to adult heroism around children. Over the summer, I worked in a Montessori Daycare as a Toddler Teacher, and I noticed that I was less fearful and more motivated while I was around my students. My discomforts and phobias were annulled if they got in the way of providing the best care for my students, and I wanted to explore this correlation. 

Based on the availability of data sets, I adapted my question about adult heroism around children to make it more applicable to the greater public sector. The question I tackled is, “Do police respond better when victims are younger?” Aware of the subjectivity of the word "better," in this research, a "better" police response is determined by victim count---people who were injured or killed. The lower the victim count, the better the police response. Due to victim confidentiality and the limited information of my data set, individual victim ages could not be determined. Therefore, victim age is assumed to be correlated with the shooting location. More specifically, I segmented the data set into two groups: "School Shooting" and "Non-School Shooting." This data is analyzed with the foundational understanding that school shootings have younger victims.


My hypothesis is that police will respond better when there are younger victims---that is, school shootings will have fewer victims than non-school shootings. 


The reason why I am making this hypothesis is rooted in the theory of reproductive value. As I've learned in the Human Nature course taught by Joe Henrich, we are evolutionarily hardwired to grieve the most for people closest to the age of prime reproduction who have the most reproductive time left. For example, this theory posits that we'd grieve more for an 11 year old than we would a 30 year old. With this rationale, police would act the most preventatively for people with the most reproductive value, aka youth. Because of this, I hypothesize that police would allow a smaller amount of victims for shootings that happen at schools, a place they would know to have children. My explanatory variable is shooting setting (school or non-school). My outcome variable is the number of victims. If I find that there are significantly fewer victims on average for school shootings, it will confirm my hypothesis.

This problem is important to investigate to determine if there's implicit bias acting in a policing system that is supposed to be egalitarian.

Link to the data I am using: https://www.kaggle.com/datasets/zusmani/us-mass-shootings-last-50-years


## Data


```{r}
# Loading Data
library(tidyverse)
library(janitor)

shootings <- read_csv("data/Mass Shootings Dataset Ver 5.csv", show_col_types = FALSE) |>
  janitor::clean_names()


# Dependent Variable Visualization
ggplot(data = shootings,
       mapping = aes(x = s_number,
                     y = total_victims)) +
  geom_point() +
  labs(x = "Shooting Recency Number (1 being most recent in 2017)",
       y = "Victim Count",
       title = "Mass Shooting Victims")

```
This is a cross-sectional observational study, meaning that we will be comparing the outcome variable (victim count) between our two groups (school shootings and non-school shootings) at one point in time. This data is aggregated through the independent research of Zeeshan-ul-hassan Usmani, Ph.D. from the start of 1966 to the end of 2017. After removing NA's, there are 261 entries in this data set. This data came primarily  from Mother Jones, which is an open-source database that uses collective expert effort to update the victim counts and other descriptive information about the United States' mass shootings as they occur. Based on this data set, the criteria for a shooting to be categorized as a "mass shooting," is having 3 or more victims, "exclud[ing] shootings stemming from more conventionally motivated crimes such as armed robbery or gang violence" (Follman et al). 

A dot plot of victim count as a function of recency (shooting number) for all 261 eligible shootings is shown above. The x-axis shows recency; the bigger the shooting number, the longer ago it occurred. Notably, there is an outlier at the left end of the graph, which condenses the data.

### Comparing the Centers of Shooting-Type Distributions

```{r}
# `shooting_type` Variable
shootings <- shootings |> 
  mutate(shooting_type = if_else(str_detect(incident_area, "School") == TRUE,
         "School Shooting",
         "Non-School Shooting")) |>
  filter(is.na(shooting_type) == FALSE)


# Average Total Victim Table
avg_victim_count <- shootings |>
  group_by(shooting_type) |>
  summarize(avg_total_victims = mean(total_victims, na.rm = TRUE))

knitr::kable(avg_victim_count, digits = 3,
             col.names = c("Shooting Type",
                           "Average Total Victims"))

```

I created a new variable called `shooting_type` that took on the label "School Shooting" if the `str_detect` function found the word "school" in the shooting's `incident_area` column. All other shootings received the label  "Non-School Shooting." I then grouped by `shooting_type` and found the average number of victims for each type of shooting. 

Based on this table, we see that the average total victims for non-school shootings is about one person greater than school shootings, which supports my hypothesis that police respond better when there are younger victims. Of course, it does not confirm the hypothesis because lots of things can contribute to victim count that are out of the police's control, such as type of gun used and the density of people in one location. 




```{r}
# Median Total Victim Table
median_victim_count <- shootings |>
  group_by(shooting_type) |>
  summarize(median_total_victims = median(total_victims, na.rm = TRUE))

knitr::kable(median_victim_count, digits = 3,
              col.names = c("Shooting Type",
                            "Median Total Victims"))
```
Since the Las Vegas Shooting in 2017 is an immense outlier---seen in the dotplot---a more accurate representation of the victim counts may be to find the median of each shooting type. Still, there is typically a greater number of victims in non-school shootings.


## Results


```{r}
# Victim Count Comparison Boxplots
library(ggplot2)

ggplot(data = shootings, aes(x = shooting_type, y = total_victims)) +
  geom_boxplot() +
  scale_y_log10() +
  labs(x = "Shooting Type",
       y = "Total Number of Victims")

```
Since we concluded that the medians better control for the outlier, this boxplot represents the median victim count in context with the other quartiles for each shooting type. Comparing the minimum and maximum victim counts, both shooting types seem to have similar minimums, but school shootings have a greater maximum, ignoring outliers. There are many more outliers in non-school shootings, which suggests that that shooting type has more spread/variation.





### Regression

```{r}
# Non-School Shooting Regression
shootings |>
  filter(shooting_type == "Non-School Shooting") |>
  ggplot(aes(x = s_number, y = total_victims)) +
  geom_point(color = "steelblue1") +
  labs(
  x = "Shooting Recency Number",
  y = "Total Number of Victims",
  title = "Non-School Shooting Victim Count Regression"
  ) +
  geom_smooth(method = "lm", se = FALSE, color = "indianred1", size = 1.5)
```
Here, we have the regression graph for non-school shootings. Again, the outlier condenses all other data points. The line of best fit is nearly flat, but slightly tilted downward reading left to right. This signifies a weak negative correlation between shooting recency number and total number of victims. This means that a higher recency number coorelates to a lower victim number. High recency numbers represent shootings that occured further in the past, so this line shows a trend for fewer victims in past non-school shootings and an increasing number of non-school shooting victims in the present day. 

Though not directly related to my adult heroism around children hypothesis, this is still important to note that mass shooting are on the rise, hinting at another dire issue.


```{r}
# School Shooting Regression
shootings |>
  filter(shooting_type == "School Shooting") |>
  ggplot(aes(x = s_number, y = total_victims)) +
  geom_point(color = "steelblue1") +
  labs(
  x = "Shooting Recency Number",
  y = "Total Number of Victims",
  title = "School Shooting Victim Count Regression"
  ) +
  geom_smooth(method = "lm", se = FALSE, color = "indianred1", size = 1.5)
```
Here, we have the regression graph for school shootings. Without the outlier that was present in the non-school shooting data, this school shooting data is more spread out within this range of victim totals. Again, the line of best fit is slightly tilted downward reading left to right. This signifies a somewhat stronger, yet still weak negative correlation between shooting recency number and total number of victims. This means that a higher recency number coorelates to a lower victim number. High recency numbers represent shootings that occured further in the past, so this line shows a trend for fewer victims in past school shootings and an increasing number of school shooting victims in the present day. Notably, many of the data points are aggreated toward the lower, less recent end of the graph, signifying that school shootings happened more frequently in years before 2017. Though, it is important to note that for the least recent shootings, there were often years before another shooting would occur, so the recency number is not a constant value. For context, a shooting recency number of 300 comes from 1988, 250 comes from 1999, and 200 comes from 2009.



```{r}
# Regression Table
model <- lm(total_victims~ shooting_type, data = shootings)
summary(model)
var_labels <- c(
  "(Intercept)" = "Intercept (Non-School Shootings)",
  "shooting_typeSchool Shooting" = "School Shootings")
modelsummary::modelsummary(model,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           coef_map = var_labels,
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))

```

This regression table indicates that school shootings have a 1.196 person lower average victim count (so 9.888 victims) compared to non-school shootings, which averaged 11.084 victims (seen as the "Intercept" coefficient). This difference is not statistically significant at the 0.05 alpha level because this p-value is 0.859, which is much higher than alpha, so any observed difference could be caused by random chance. Because of this, we would fail to reject a null hypothesis of no difference between school shooting and non-school shooting victim counts. Our r^2 value is 0, which means that basically none of the variation in the victim counts can be explained by the shooting type. This coefficient most certainly does not show a causal affect; there's basically no relation between the two based on the data.


## Conclusion
After analyzing the data, we found that when comparing the means and medians of victim counts in school shootings vs. non-school shootings, it appears that school shootings tend to have fewer victims, which would support my hypothesis that police respond better when there are younger victims. However, when looking at the statistical significance of this data granted by the regression table, we see that this is insignificant data, and we therefore cannot make any conclusions about our hypothesis. 

To discuss a few limitations and confounders in this data analysis: 

There were lots of NAs that caused whole shootings to be removed from the data set, shrinking our sample size. This data set also did not include the dates in a manner which could have easily been configured into an x-axis of time. Because of this, I was forced to use the s_number, which was conveniently correlated with the recency of the shooting, but the time that separated one s_number from the next varied, so it wasn't the best variable to have on an axis since that variation in time was not shown on the axis.
 
Though my initial question pondered whether police respond better to shootings with younger victims, I had no way of knowing the victim ages in these shootings. I got around this roadblock by making the `shooting_type` variable and selecting shootings with "school" in the description, but doing this might not have gotten all the shootings that happened at schools, nor all the shootings that involved young victims.

Also, as mentioned above, victim count doesn't entirely explain the quality of police response because there are several factors that could affect victim count, such as type of gun used and density of the victims.

Lastly, this question of adult heroism around children would have been better answered by analyzing police response times and investigating a correlation between response time and victim ages. If I had more time to scour the internet, I could possibly find a data set from a police department or 911 call center that may provide me with information that is more useful in answering my question. If that type of data set doesn't exist yet, I could purchase response time information from police departments if I had enough money.  



### Works Cited

Follman, Mark, et al., editors. "A Guide to Mass Shootings in America." Mother Jones, Mother Jones
  and the Foundation for National Progress, 6 Dec. 2023,
  www.motherjones.com/politics/2012/07/mass-shootings-map/. Accessed 13 Dec. 2023. 
  
Henrich, Joseph, narrator. Lecture. 2023. Canvas,
  matterhorn.dce.harvard.edu/engage/player/watch.html?id=c2106090-cd20-18e0-39f4-e4af3e8d7bb1.         Accessed 13 Dec. 2023. 
  
Usmaniv, Zeeshan-ul-hassan. "US Mass Shootings." Kaggle, 2017,
  www.kaggle.com/datasets/zusmani/us-mass-shootings-last-50-years. Accessed 13 Dec. 2023. Table. 